---
title: Data Processing
---

A step-by-step proof of the data pipeline. We discover issues in the raw data and fix them immediately.

## 0. Setup

```{r}
library(dplyr)
library(readxl)
library(tidyr)
library(stringr)
library(nanoparquet)
```

## 1. Load and show raw

```{r}
df_pos_raw <- read_excel("data/positive-cases-Jennys-data-Edited.xlsx") %>%
  mutate(row_number = row_number())

df_neg_raw <- read_excel("data/negative-cases-ANR-Data.xlsx") %>%
  mutate(row_number = row_number())
```

Looking at raw data. The ID column (`Alias`) in the negative set contains merged cells (read as NAs).

```{r}
df_pos_raw %>% select(row_number, Alias) %>% head(27)
```

```{r}
df_pos_raw %>% select(row_number, Alias) %>% tail(27)
```

```{r}
df_neg_raw %>% select(row_number, Alias) %>% head(27)
```

```{r}
df_neg_raw %>% select(row_number, Alias) %>% tail(27)
```

```{r}
print(sprintf("Positive dimensions: %d x %d", nrow(df_pos_raw), ncol(df_pos_raw)))
print(sprintf("Negative dimensions: %d x %d", nrow(df_neg_raw), ncol(df_neg_raw)))
```

## 2. Empty rows

We filter completely empty rows before any other processing.

```{r}
# Positive empty rows
pos_empty_rows <- df_pos_raw %>% filter(if_all(everything(), is.na))
```

```{r}
pos_empty_rows
```

```{r}
# Negative empty rows
neg_empty_rows <- df_neg_raw %>% filter(if_all(everything(), is.na))
```

```{r}
neg_empty_rows
```

```{r}
# Filter and print count
rows_before_pos <- nrow(df_pos_raw)
df_pos_raw <- df_pos_raw %>% filter(!if_all(everything(), is.na))
rows_after_pos <- nrow(df_pos_raw)

rows_before_neg <- nrow(df_neg_raw)
df_neg_raw <- df_neg_raw %>% filter(!if_all(everything(), is.na))
rows_after_neg <- nrow(df_neg_raw)

print(sprintf("Positive empty rows dropped: %d", rows_before_pos - rows_after_pos))
print(sprintf("Negative empty rows dropped: %d", rows_before_neg - rows_after_neg))
```

## 3. Forward-fill the ID

**LOADBEARING** — `observation` numbering derives from this, and every first-look analysis derives from `observation == 1`.
Consumed by: `01` first-look panel, `02` first-look GLM, `03` first-look models.

The `Alias` column in the negative dataset uses merged cells, which appear as NAs. We propagate the ID downward.

```{r}
# Positive view before
before_head_pos <- df_pos_raw %>% select(row_number, Alias) %>% head(27)
before_tail_pos <- df_pos_raw %>% select(row_number, Alias) %>% tail(27)
```

```{r}
# Positive mutation
na_before_pos <- sum(is.na(df_pos_raw$Alias))
df_pos_raw <- df_pos_raw %>% fill(Alias, .direction = "down")
na_after_pos <- sum(is.na(df_pos_raw$Alias))

after_head_pos <- df_pos_raw %>% select(row_number, Alias) %>% head(27)
after_tail_pos <- df_pos_raw %>% select(row_number, Alias) %>% tail(27)

print(sprintf("Positive Alias NAs filled: %d of %d", na_before_pos - na_after_pos, na_before_pos))
```

```{r}
before_head_pos
```

```{r}
after_head_pos
```

```{r}
before_tail_pos
```

```{r}
after_tail_pos
```

```{r}
# Negative view before
before_head_neg <- df_neg_raw %>% select(row_number, Alias) %>% head(27)
before_tail_neg <- df_neg_raw %>% select(row_number, Alias) %>% tail(27)
```

```{r}
# Negative mutation
na_before_neg <- sum(is.na(df_neg_raw$Alias))
df_neg_raw <- df_neg_raw %>% fill(Alias, .direction = "down")
na_after_neg <- sum(is.na(df_neg_raw$Alias))

after_head_neg <- df_neg_raw %>% select(row_number, Alias) %>% head(27)
after_tail_neg <- df_neg_raw %>% select(row_number, Alias) %>% tail(27)

print(sprintf("Negative Alias NAs filled: %d of %d", na_before_neg - na_after_neg, na_before_neg))
```

```{r}
before_head_neg
```

```{r}
after_head_neg
```

```{r}
before_tail_neg
```

```{r}
after_tail_neg
```

## 4. Observation numbering

We assign a sequential `observation` number within each ID group.

```{r}
# Verify row order is observation order (by checking if Alias is sorted)
# We sort explicitly to be safe and transparent.
df_pos_raw <- df_pos_raw %>% arrange(Alias, `Date/Time`)
df_neg_raw <- df_neg_raw %>% arrange(Alias, `Date/Time`)
```

```{r}
df_pos_raw <- df_pos_raw %>% group_by(Alias) %>% mutate(observation = row_number()) %>% ungroup()
df_neg_raw <- df_neg_raw %>% group_by(Alias) %>% mutate(observation = row_number()) %>% ungroup()
```

Showing a slice spanning ID boundaries.

```{r}
df_neg_raw %>%
  select(row_number, Alias, observation) %>%
  slice(10:20)
```

Print distribution of observations per patient.

```{r}
pos_obs_dist <- df_pos_raw %>% count(Alias, name = "n")
neg_obs_dist <- df_neg_raw %>% count(Alias, name = "n")

print("Positive Observation Distribution:")
print(summary(pos_obs_dist$n))

print("Negative Observation Distribution:")
print(summary(neg_obs_dist$n))
```

```{r}
min_pos <- min(pos_obs_dist$n)
min_neg <- min(neg_obs_dist$n)
print(sprintf("Min observation count - Positive: %d, Negative: %d", min_pos, min_neg))
if(min_pos < 1 || min_neg < 1) stop("Minimum observation must be 1")
```

## 5. Column names

**LOADBEARING** — the cleaner permits hyphens, and a hyphenated column name parses as subtraction inside a formula. `02` backticks its formulas for this reason.
Consumed by: `02` model formulas.

We normalize column names to lowercase, replacing spaces and slashes with underscores, and removing non-alphanumeric characters (except underscores and hyphens).

```{r}
clean_name <- function(name) {
  name <- trimws(name)
  name <- gsub("/", "_", name)
  name <- gsub(" ", "_", name)
  name <- gsub("[^a-zA-Z0-9_-]", "", name)
  return(tolower(name))
}

# Positive mapping
old_pos <- names(df_pos_raw)
new_pos <- clean_name(old_pos)
map_pos <- data.frame(old = old_pos, new = new_pos, stringsAsFactors = FALSE)

# Negative mapping
old_neg <- names(df_neg_raw)
new_neg <- clean_name(old_neg)
map_neg <- data.frame(old = old_neg, new = new_neg, stringsAsFactors = FALSE)
```

Positive column mapping:

```{r}
map_pos
```

Negative column mapping:

```{r}
map_neg
```

Audit for non-standard characters (outside `[a-z0-9_-]`).

```{r}
all_cleaned <- c(map_pos$new, map_neg$new)
weird <- all_cleaned[grepl("[^a-z0-9_-]", all_cleaned)]
if(length(weird) > 0) {
  print("Names with non-standard characters:")
  print(weird)
} else {
  print("✓ All cleaned names follow [a-z0-9_-] pattern.")
}
```

Apply the cleaning.

```{r}
names(df_pos_raw) <- new_pos
names(df_neg_raw) <- new_neg
```

## 6. Semantic alignment

We align columns that represent the same measurement but have different names.

Candidates for alignment:

```{r}
# BMI candidates
pos_bmi <- grep("bmi", names(df_pos_raw), value = TRUE)
neg_bmi <- grep("bmi", names(df_neg_raw), value = TRUE)
# O2 candidates
pos_o2 <- grep("o2", names(df_pos_raw), value = TRUE)
neg_o2 <- grep("o2", names(df_neg_raw), value = TRUE)
```

Showing candidates side by side.

```{r}
# Positive BMI/O2 head
df_pos_raw %>% select(all_of(c(pos_bmi, pos_o2))) %>% head(27)
```

```{r}
# Negative BMI/O2 head
df_neg_raw %>% select(all_of(c(neg_bmi, neg_o2))) %>% head(27)
```

Applying renames.

```{r}
semantic_map <- c(
  "bmicalc" = "bmi",
  "o2sat" = "o2_sat"
)

# Positive renames
pos_to_rename <- names(semantic_map)[names(semantic_map) %in% names(df_pos_raw)]
df_pos_raw <- df_pos_raw %>%
  rename_with(~ semantic_map[.x], .cols = all_of(pos_to_rename))

# Negative renames
neg_to_rename <- names(semantic_map)[names(semantic_map) %in% names(df_neg_raw)]
df_neg_raw <- df_neg_raw %>%
  rename_with(~ semantic_map[.x], .cols = all_of(neg_to_rename))

print(sprintf("Positive renames fired: %s", paste(pos_to_rename, collapse = ", ")))
print(sprintf("Negative renames fired: %s", paste(neg_to_rename, collapse = ", ")))
```

## 7. Labels

Assign 1 to positives, 0 to negatives.

```{r}
df_pos_raw <- df_pos_raw %>% mutate(progression_to_death = 1)
df_neg_raw <- df_neg_raw %>% mutate(progression_to_death = 0)
```

```{r}
df_pos_raw %>% count(progression_to_death)
```

```{r}
df_neg_raw %>% count(progression_to_death)
```

## 8. Type alignment

We identify columns present in both workbooks with differing types.

```{r}
common_cols <- intersect(names(df_pos_raw), names(df_neg_raw))
mismatches <- lapply(common_cols, function(col) {
  type_pos <- class(df_pos_raw[[col]])[1]
  type_neg <- class(df_neg_raw[[col]])[1]
  if(type_pos != type_neg) {
    return(data.frame(
      column = col,
      pos_type = type_pos,
      neg_type = type_neg,
      pos_sample = paste(head(df_pos_raw[[col]], 3), collapse = ", "),
      neg_sample = paste(head(df_neg_raw[[col]], 3), collapse = ", "),
      stringsAsFactors = FALSE
    ))
  }
  return(NULL)
})

mismatch_df <- bind_rows(mismatches)
```

Type mismatches:

```{r}
mismatch_df
```

### Datetime alignment

```{r}
# All-null columns in negative dataset: set to typed NA
datetime_nulls <- c("date_time_of_declaration_tod", "date_time_of_pea_asystole", "date_time_of_perfusion")
for(col in datetime_nulls) {
  df_neg_raw[[col]] <- as.POSIXct(NA)
}
```

```{r}
# Parsing date_time_sbp__90
na_before <- sum(is.na(df_neg_raw$date_time_sbp__90))
df_neg_raw$date_time_sbp__90 <- as.POSIXct(df_neg_raw$date_time_sbp__90, format = "%Y-%m-%d %H:%M:%S")
na_after <- sum(is.na(df_neg_raw$date_time_sbp__90))
print(sprintf("date_time_sbp__90 NAs: %d -> %d", na_before, na_after))
if(na_after > na_before) stop("NA count increased during date parsing")
```

### Integer alignment

```{r}
int_nulls <- c("dcd_nrp_total_pump_time", "extubation_to_perfusion_warm_ischemic_time",
               "sbp90_to_declaration", "tod_to_perfusion", "warm_ischemic_time_agonal_phase_to_cooling")

for(col in int_nulls) {
  na_before <- sum(is.na(df_neg_raw[[col]]))
  df_neg_raw[[col]] <- as.integer(df_neg_raw[[col]])
  na_after <- sum(is.na(df_neg_raw[[col]]))
  print(sprintf("%s NAs: %d -> %d", col, na_before, na_after))
  if(na_after > na_before) stop(sprintf("NA count increased during int cast for %s", col))
}
```

## 9. Namespace the patient IDs

**LOADBEARING** — both workbooks number patients from 1. Without the prefix, `pos_1` and `neg_1` collapse into one group, which corrupts patient counts, degrades group-aware stratification, and makes the group structure meaningless.
Consumed by: `01` patient counts, `02` patient-level aggregation, `03` group splits.

```{r}
pos_ids_before <- n_distinct(df_pos_raw$alias)
neg_ids_before <- n_distinct(df_neg_raw$alias)

df_pos_raw <- df_pos_raw %>% mutate(alias = paste0("pos_", alias))
df_neg_raw <- df_neg_raw %>% mutate(alias = paste0("neg_", alias))

pos_ids_after <- n_distinct(df_pos_raw$alias)
neg_ids_after <- n_distinct(df_neg_raw$alias)
```

```{r}
print(sprintf("Unique IDs - Pos: %d -> %d", pos_ids_before, pos_ids_after))
print(sprintf("Unique IDs - Neg: %d -> %d", neg_ids_before, neg_ids_after))
```

```{r}
df_pos_raw %>% select(alias) %>% head(27)
```

```{r}
df_neg_raw %>% select(alias) %>% head(27)
```

## 10. Combine

**LOADBEARING** — a column populated in one workbook and absent or empty in the other is a perfect proxy for the label. The missingness filter in `01` is what catches these, so a reader needs to see the list here to know what that filter is doing.
Consumed by: `01` missingness filter and leak exclusion list.

```{r}
df_all <- bind_rows(df_pos_raw, df_neg_raw) %>% 
  mutate(alias_filled = alias)
```

Asymmetric columns:

```{r}
cols_pos <- names(df_pos_raw)
cols_neg <- names(df_neg_raw)

pos_only <- setdiff(cols_pos, cols_neg)
neg_only <- setdiff(cols_neg, cols_pos)
both <- intersect(cols_pos, cols_neg)

print(sprintf("Columns in both: %d", length(both)))
print(sprintf("Columns in Positives only: %d", length(pos_only)))
print(sprintf("Columns in Negatives only: %d", length(neg_only)))
```

```{r}
pos_only
```

```{r}
neg_only
```

```{r}
print(sprintf("Final dimensions: %d x %d", nrow(df_all), ncol(df_all)))
```

## 11. Drop the scaffolding

We remove `row_number` before writing.

```{r}
df_all <- df_all %>% select(-row_number)
```

```{r}
names(df_all)
```

## 12. Write

```{r}
nanoparquet::write_parquet(df_all, "data/processed/combined-dataset.parquet")
```

```{r}
print(sprintf("Saved to data/processed/combined-dataset.parquet"))
print(sprintf("Dimensions: %d x %d", nrow(df_all), ncol(df_all)))
```

Final Summary:

```{r}
df_all %>%
  group_by(progression_to_death) %>%
  summarise(
    rows = n(),
    patients = n_distinct(alias)
  )
```

```{r}
df_all %>%
  group_by(progression_to_death, alias) %>%
  summarise(obs = n(), .groups = "drop") %>%
  group_by(progression_to_death) %>%
  summarise(
    avg_obs = mean(obs),
    max_obs = max(obs),
    min_obs = min(obs)
  )
```
