---
title: Data Exploration
---

Exploratory analysis with detailed missingness patterns.

## 0. Setup

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(nanoparquet)
```

```{r}
dir.create("output", showWarnings = FALSE)
df <- nanoparquet::read_parquet("data/processed/combined-dataset.parquet")
```

## 1. Class and patient distributions

Rows per class:

```{r}
df %>% count(progression_to_death)
```

Patients per class and observations per patient.

```{r}
df %>% 
  group_by(progression_to_death, alias_filled) %>% 
  summarise(obs = n(), .groups = "drop") %>% 
  group_by(progression_to_death) %>% 
  summarise(
    count = n(),
    avg_obs = mean(obs),
    max_obs = max(obs),
    min_obs = min(obs)
  )
```

Observations-per-patient distribution. This variance informs the equal-weighting limitation in the patient-level GLM.

```{r}
df %>% 
  group_by(progression_to_death, alias_filled) %>% 
  summarise(obs = n(), .groups = "drop") %>% 
  count(progression_to_death, obs) %>% 
  arrange(progression_to_death, obs)
```

## 2. Within-patient stability

Quantify variance within patients relative to population variance.

```{r}
# Identify numeric features (excluding label and ID)
numeric_cols <- df %>% 
  select(where(is.numeric)) %>% 
  select(-any_of(c("progression_to_death", "observation", "alias", "alias_filled"))) %>% 
  names()

stability_results <- lapply(numeric_cols, function(col) {
  global_sd <- sd(df[[col]], na.rm = TRUE)
  if(is.na(global_sd) || global_sd == 0) return(NULL)
  
  within_sd <- df %>% 
    group_by(alias_filled) %>% 
    summarise(s = sd(!!sym(col), na.rm = TRUE), .groups = "drop") %>% 
    filter(!is.na(s)) %>% 
    pull(s) %>% 
    mean()
  
  if(is.na(within_sd)) return(NULL)
  
  data.frame(feature = col, ratio = within_sd / global_sd, stringsAsFactors = FALSE)
}) %>% bind_rows() %>% arrange(ratio)
```

```{r}
stability_results %>% 
  mutate(status = if_else(ratio < 0.3, "Stable", "Volatile"))
```

```{r}
print(sprintf("Average Stability Ratio: %.3f", mean(stability_results$ratio)))
```

**Interpretation:** The Stability Ratio ($\frac{\text{mean}(\sigma_{\text{patient}})}{\sigma_{\text{global}}}$) justifies the "First Look Only" design. A ratio < 0.3 suggests repeated observations are largely redundant. 

Note that `global_sd` is computed on pooled data across both classes, meaning it includes between-class variance, which makes the ratio flatter than a true within-versus-between patient comparison.

## 3. Missingness

Missing count and percent per column.

```{r}
missing_stats <- df %>% 
  summarise(across(everything(), ~sum(is.na(.x)))) %>% 
  pivot_longer(everything(), names_to = "feature", values_to = "null_count") %>% 
  mutate(null_pct = null_count / nrow(df) * 100) %>% 
  arrange(desc(null_pct))
```

```{r}
missing_stats
```

**LOADBEARING** — A column populated in one workbook and empty in the other is a perfect label proxy. This table is where such leaks are first visible.
Consumed by: the missingness filter, and the leak discussion in `00`'s concat provenance.

```{r}
missing_by_class <- df %>% 
  group_by(progression_to_death) %>% 
  summarise(across(everything(), ~sum(is.na(.x)))) %>% 
  pivot_longer(-progression_to_death, names_to = "feature", values_to = "null_count")

class_totals <- df %>% count(progression_to_death)

missing_by_class <- missing_by_class %>% 
  left_join(class_totals, by = "progression_to_death") %>% 
  mutate(null_pct = null_count / n * 100) %>% 
  filter(null_count > 0) %>% 
  select(feature, progression_to_death, null_count, null_pct) %>% 
  pivot_wider(names_from = progression_to_death, values_from = c(null_count, null_pct), names_sep = "_")
```

```{r}
missing_by_class
```

Missingness Heatmap.

```{r}
# Downsample for heatmap
set.seed(8675309)
heatmap_df <- df %>% 
  select(all_of(numeric_cols)) %>% 
  mutate(across(everything(), ~as.factor(as.integer(is.na(.x))))) %>% 
  slice_sample(n = min(nrow(.), 1000))

# Note: ggplot2 heatmap usually requires long format
heatmap_long <- heatmap_df %>% 
  mutate(row_id = row_number()) %>% 
  pivot_longer(-row_id) %>% 
  mutate(value = as.factor(value))

ggplot(heatmap_long, aes(x = name, y = row_id, fill = value)) +
  geom_tile() +
  scale_fill_manual(values = c("white", "black"), labels = c("Present", "Missing"), name = "Status") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(title = "Missingness Heatmap", x = "Feature", y = "Row Index")
```

Distribution of missing features per row.

```{r}
df %>% 
  select(all_of(numeric_cols)) %>% 
  mutate(across(everything(), ~as.integer(is.na(.x)))) %>% 
  rowSums() %>% 
  table() %>% 
  as.data.frame() %>% 
  setNames(c("missing_count", "row_count"))
```

## 4. Descriptive separation by outcome

We check if modelling results are visible in raw data using **first observations only** (one row per patient).

```{r}
df_first <- df %>% 
  filter(observation == 1)
```

Numeric separation (Wilcoxon rank-sum).

```{r}
numeric_sep <- lapply(numeric_cols, function(col) {
  pos_vals <- df_first %>% filter(progression_to_death == 1) %>% pull(!!sym(col))
  neg_vals <- df_first %>% filter(progression_to_death == 0) %>% pull(!!sym(col))
  
  if(length(pos_vals) < 1 || length(neg_vals) < 1) return(NULL)
  
  # Use a tryCatch to avoid crashing on degenerate data (e.g. all NAs)
  test_res <- tryCatch({
    wilcox.test(pos_vals, neg_vals)
  }, error = function(e) NULL)
  
  if(is.null(test_res)) return(NULL)
  
  data.frame(
    feature = col,
    pos_median = median(pos_vals, na.rm = TRUE),
    neg_median = median(neg_vals, na.rm = TRUE),
    p_value = test_res$p.value,
    stringsAsFactors = FALSE
  )
}) %>% bind_rows() %>% arrange(p_value)
```

```{r}
numeric_sep
```

Categorical separation (Fisher/Chi-sq).

```{r}
categorical_cols <- df %>% 
  select(where(is.character)) %>% 
  select(-any_of(c("alias", "alias_filled"))) %>% 
  names()

cat_sep <- lapply(categorical_cols, function(col) {
  tab <- table(df_first[[col]], df_first$progression_to_death)
  if(nrow(tab) < 2 || ncol(tab) < 2) return(NULL)
  
  test_res <- tryCatch({
    if(any(tab < 5)) fisher.test(tab) else chisq.test(tab)
  }, error = function(e) NULL)
  
  if(is.null(test_res)) return(NULL)
  
  data.frame(
    feature = col,
    p_value = test_res$p.value,
    stringsAsFactors = FALSE
  )
}) %>% bind_rows() %>% arrange(p_value)
```

```{r}
cat_sep
```

Top 9 features by separation (visual).

```{r}
# Pick top 9 numeric features by p-value
top_9 <- head(numeric_sep$feature, 9)

df_first %>% 
  select(progression_to_death, all_of(top_9)) %>% 
  pivot_longer(-progression_to_death) %>% 
  ggplot(aes(x = factor(progression_to_death), y = value)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.5) +
  facet_wrap(~name, scales = "free") +
  labs(x = "Class (0=Neg, 1=Pos)", y = "Value") +
  theme_minimal()
```

These tests rank features for plotting and are not an inferential claim. Separation here is often weaker than final model AUC because the model combines multiple features.

## 5. Feature filters

### Exclusions

**LOADBEARING** — These are post-outcome variables; including any of them makes the ML results meaningless.
Consumed by: `03` model pipeline.

```{r}
leak_justifications <- tibble(
  feature = c(
    "dcd_nrp_total_pump_time",
    "extubation_to_perfusion_warm_ischemic_time",
    "tod_to_perfusion",
    "sbp90_to_declaration",
    "warm_ischemic_time_agonal_phase_to_cooling",
    "did_patient_expire_within_timeframe"
  ),
  reason = c(
    "Post-outcome: pump time recorded during DCD/NRP",
    "Post-outcome: time from extubation to perfusion",
    "Post-outcome: time from declaration to perfusion",
    "Post-outcome: time from SBP<90 to declaration",
    "Post-outcome: agonal phase duration",
    "Direct proxy: patient expiration status"
  )
)

id_excludes <- c("alias", "alias_filled", "observation")
all_excludes <- c(leak_justifications$feature, id_excludes)
```

```{r}
leak_justifications
```

```{r}
candidates <- names(df)
before_exc <- candidates
exc_dropped_table <- tibble(
  feature = all_excludes,
  reason = if_else(all_excludes %in% leak_justifications$feature, 
                   leak_justifications$reason[match(all_excludes, leak_justifications$feature)], 
                   "Manual/ID Exclusion")
)
candidates <- setdiff(before_exc, exc_dropped_table$feature)
print(sprintf("Exclusions: %d in, %d out, %d dropped", 
              length(before_exc), length(candidates), nrow(exc_dropped_table)))
```

```{r}
exc_dropped_table
```

### Missingness

Threshold: 10%.

```{r}
before_miss <- candidates
missing_vals <- df %>% 
  summarise(across(all_of(candidates), ~sum(is.na(.x)) / n() * 100)) %>% 
  pivot_longer(everything(), names_to = "feature", values_to = "pct")

miss_dropped_table <- missing_vals %>% 
  filter(pct > 10) %>% 
  select(feature, reason = pct)

candidates <- setdiff(before_miss, miss_dropped_table$feature)
print(sprintf("Missingness: %d in, %d out, %d dropped", 
              length(before_miss), length(candidates), nrow(miss_dropped_table)))
```

```{r}
miss_dropped_table
```

### Variance

Threshold: 0.01 (raw unscaled variance). This is unit-dependent and primarily catches degenerate columns.

```{r}
before_var <- candidates
numeric_candidates <- candidates[sapply(candidates, function(c) is.numeric(df[[c]]))]

var_vals <- df %>% 
  summarise(across(all_of(numeric_candidates), var)) %>% 
  pivot_longer(everything(), names_to = "feature", values_to = "variance")

var_dropped_table <- var_vals %>% 
  filter(variance < 0.01 | is.na(variance)) %>% 
  mutate(reason = sprintf("%.4f", variance)) %>%
  select(feature, reason)

candidates <- setdiff(before_var, var_dropped_table$feature)
print(sprintf("Variance: %d in, %d out, %d dropped", 
              length(before_var), length(candidates), nrow(var_dropped_table)))
```

```{r}
var_dropped_table
```

## 6. Correlation

We report highly correlated numeric features (|r| > 0.9). This section reports pairs but drops nothing.

```{r}
numeric_final <- candidates[sapply(candidates, function(c) is.numeric(df[[c]]))]
cor_matrix <- cor(df %>% select(all_of(numeric_final)), use = "pairwise.complete.obs")

cor_pairs <- as.data.frame(as.table(cor_matrix)) %>% 
  filter(Var1 != Var2) %>% 
  filter(abs(Freq) > 0.9) %>% 
  distinct(Var1, Var2, .keep_all = TRUE) %>% 
  rename(feature1 = Var1, feature2 = Var2, r = Freq)
```

```{r}
if(nrow(cor_pairs) > 0) {
  cor_pairs
} else {
  print("No highly correlated pairs (|r| > 0.9) found.")
}
```

## 7. Write outputs

```{r}
# Analytic dataset: Selected features + scaffolding
df_analytic <- df %>% 
  select(all_of(candidates), any_of(c("alias_filled", "observation", "progression_to_death")))

nanoparquet::write_parquet(df_analytic, "data/processed/analytic-dataset.parquet")
```

```{r}
# Selected features artifact
feature_artifact <- missing_vals %>% 
  left_join(var_vals, by = "feature") %>% 
  filter(feature %in% candidates) %>% 
  select(feature, missing_pct = pct, variance)

write.csv(feature_artifact, "output/selected-features.csv", row.names = FALSE)
```

```{r}
print(sprintf("Final feature set: %d features", length(candidates)))
cat(paste(candidates, collapse = "\n"))
```
