

# Exploratory Analysis

Exploratory analysis with detailed missingness patterns.


```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(nanoparquet)
```

```{r}
df <- nanoparquet::read_parquet("data/processed/combined-dataset.parquet")
```

## 1. Class and patient distributions

Rows per class:

```{r}
df %>%
  count(progression_to_death)
```

Visualize the patients per class and observations per patient.

```{r}
df %>%
  group_by(progression_to_death, alias_filled) %>%
  summarise(obs = n(), .groups = "drop") %>%
  group_by(progression_to_death) %>%
  summarise(
    count = n(),
    avg_obs = mean(obs),
    max_obs = max(obs),
    min_obs = min(obs)
  )
```

We visualize the distribution of observations per patient, stratified by outcome.
This checks if one class is more "dense" than the other, which could lead to temporal leakage.

```{r}
df %>%
  group_by(progression_to_death, alias_filled) %>%
  summarise(obs = n(), .groups = "drop") %>%
  ggplot(aes(x = obs, fill = factor(progression_to_death))) +
  geom_histogram(binwidth = 1, alpha = 0.4, position = "identity", color = "white") +
  theme_minimal() +
  labs(title = "Observations per Patient Distribution",
       x = "Number of Observations",
       y = "Patient Count",
       fill = "Outcome (0=Neg, 1=Pos)")
```

# 2. Within-patient stability

We quantify how much features fluctuate within a patient relative to the population.
A low ratio suggests that repeated measurements are redundant, justifying a "First Look Only" model.

```{r}
# 1. Global SDs
global_sds <- df %>%
  select(where(is.numeric), -any_of(c("progression_to_death", "observation", "alias", "alias_filled"))) %>%
  summarise(across(everything(), ~sd(.x, na.rm = TRUE))) %>%
  pivot_longer(everything(), names_to = "feature", values_to = "global_sd")

# 2. Mean of within-patient SDs
within_sds <- df %>%
  select(where(is.numeric), alias_filled, -any_of(c("progression_to_death", "observation"))) %>%
  pivot_longer(-alias_filled, names_to = "feature", values_to = "val") %>%
  group_by(feature, alias_filled) %>%
  summarise(s = sd(val, na.rm = TRUE), .groups = "drop") %>%
  group_by(feature) %>%
  summarise(within_sd = mean(s, na.rm = TRUE), .groups = "drop")

# 3. Combine and calculate ratio
stability_results <- global_sds %>%
  inner_join(within_sds, by = "feature") %>%
  mutate(ratio = within_sd / global_sd) %>%
  filter(!is.na(ratio), is.finite(ratio)) %>%
  arrange(ratio)

# Store for later use in heatmap
numeric_cols <- stability_results$feature
```

```{r}
stability_results <- stability_results %>%
  mutate(status = if_else(ratio < 0.3, "Stable", "Volatile"))

stability_results %>%
  print(n=Inf)
```

```{r}
print(sprintf("Average Stability Ratio: %.3f",
              mean(stability_results$ratio, na.rm = TRUE)))
```

## 3. Missingness

We analyze missing count and percent per column.

```{r}
missing_stats <- df %>%
  summarise(across(everything(), ~sum(is.na(.x)))) %>%
  pivot_longer(everything(), names_to = "feature", values_to = "null_count") %>%
  mutate(null_pct = null_count / nrow(df) * 100) %>%
  arrange(desc(null_pct))

missing_stats
```

**LOADBEARING** — A column populated in one workbook and empty in the other is a perfect label proxy. This table is where such leaks are first visible.
Consumed by: the missingness filter, and the leak discussion in `00`'s concat provenance.

```{r}
missing_by_class <- df %>%
  group_by(progression_to_death) %>%
  summarise(across(everything(), ~sum(is.na(.x)))) %>%
  pivot_longer(-progression_to_death, names_to = "feature", values_to = "null_count")

class_totals <- df %>%
  count(progression_to_death)

missing_by_class <- missing_by_class %>%
  left_join(class_totals, by = "progression_to_death") %>%
  mutate(null_pct = null_count / n * 100) %>%
  filter(null_count > 0) %>%
  select(feature, progression_to_death, null_count, null_pct) %>%
  pivot_wider(names_from = progression_to_death, values_from = c(null_count, null_pct), names_sep = "_")

missing_by_class
```

We visualize missingness using a heatmap.

```{r}
heatmap_df <- df %>%
  select(all_of(numeric_cols)) %>%
  mutate(across(everything(), ~as.factor(as.integer(is.na(.x)))))

# Note: ggplot2 heatmap usually requires long format
heatmap_long <- heatmap_df %>%
  mutate(row_id = row_number()) %>%
  pivot_longer(-row_id) %>%
  mutate(value = as.factor(value))

ggplot(heatmap_long, aes(x = name, y = row_id, fill = value)) +
  geom_tile() +
  scale_fill_manual(values = c("white", "black"), labels = c("Present", "Missing"), name = "Status") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(title = "Missingness Heatmap", x = "Feature", y = "Row Index")
```

Distribution of missing features per row.

```{r}
# Exclude ID and label columns to ensure we only count feature missingness
df %>%
  select(-any_of(c("progression_to_death", "observation", "alias", "alias_filled"))) %>%
  mutate(across(everything(), ~as.integer(is.na(.x)))) %>%
  rowSums() %>%
  table() %>%
  as.data.frame() %>%
  setNames(c("missing_count", "row_count"))
```

## 4. Descriptive separation by outcome

We check if modelling results are visible in raw data using **first observations only** (one row per patient).

```{r}
df_first <- df %>%
  filter(observation == 1)
```

Numeric separation (Wilcoxon rank-sum).

```{r}
numeric_sep <- lapply(numeric_cols, function(col) {
  pos_vals <- df_first %>% filter(progression_to_death == 1) %>% pull(!!sym(col))
  neg_vals <- df_first %>% filter(progression_to_death == 0) %>% pull(!!sym(col))

  if(length(pos_vals) < 1 || length(neg_vals) < 1) return(NULL)

  # Use a tryCatch to avoid crashing on degenerate data (e.g. all NAs)
  test_res <- tryCatch({
    wilcox.test(pos_vals, neg_vals)
  }, error = function(e) NULL)

  if(is.null(test_res)) return(NULL)

  data.frame(
    feature = col,
    pos_median = median(pos_vals, na.rm = TRUE),
    neg_median = median(neg_vals, na.rm = TRUE),
    p_value = test_res$p.value,
    stringsAsFactors = FALSE
  )
}) %>% bind_rows() %>% arrange(p_value)
```

```{r}
numeric_sep
```

Categorical separation (Fisher/Chi-sq).

```{r}
categorical_cols <- df %>%
  select(where(is.character)) %>%
  select(-any_of(c("alias", "alias_filled"))) %>%
  names()

cat_sep <- lapply(categorical_cols, function(col) {
  tab <- table(df_first[[col]], df_first$progression_to_death)
  if(nrow(tab) < 2 || ncol(tab) < 2) return(NULL)

  test_res <- tryCatch({
    if(any(tab < 5)) fisher.test(tab) else chisq.test(tab)
  }, error = function(e) NULL)

  if(is.null(test_res)) return(NULL)

  data.frame(
    feature = col,
    p_value = test_res$p.value,
    stringsAsFactors = FALSE
  )
}) %>% bind_rows() %>% arrange(p_value)
```

```{r}
cat_sep
```

Top 9 features by separation (visual).

```{r}
# Pick top 9 numeric features by p-value
top_9 <- head(numeric_sep$feature, 9)

df_first %>%
  select(progression_to_death, all_of(top_9)) %>%
  pivot_longer(-progression_to_death) %>%
  ggplot(aes(x = factor(progression_to_death), y = value)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.5) +
  facet_wrap(~name, scales = "free") +
  labs(x = "Class (0=Neg, 1=Pos)", y = "Value") +
  theme_minimal()
```

These tests rank features for plotting and are not an inferential claim. Separation here is often weaker than final model AUC because the model combines multiple features.

## 5. Feature filters

# Exclusions

**LOADBEARING** — These are post-outcome variables; including any of them makes the ML results meaningless.
Consumed by: `03` model pipeline.

```{r}
leak_justifications <- tibble(
  feature = c(
    "dcd_nrp_total_pump_time",
    "extubation_to_perfusion_warm_ischemic_time",
    "tod_to_perfusion",
    "sbp90_to_declaration",
    "warm_ischemic_time_agonal_phase_to_cooling",
    "did_patient_expire_within_timeframe"
  ),
  reason = c(
    "Post-outcome: pump time recorded during DCD/NRP",
    "Post-outcome: time from extubation to perfusion",
    "Post-outcome: time from declaration to perfusion",
    "Post-outcome: time from SBP<90 to declaration",
    "Post-outcome: agonal phase duration",
    "Direct proxy: patient expiration status"
  )
)

id_excludes <- c("alias", "alias_filled", "observation")
all_excludes <- c(leak_justifications$feature, id_excludes)

exc_dropped_table <- tibble(
  feature = all_excludes,
  reason = if_else(all_excludes %in% leak_justifications$feature,
                   leak_justifications$reason[match(all_excludes, leak_justifications$feature)],
                   "Manual/ID Exclusion")
)
```

```{r}
leak_justifications
```

```{r}
print(sprintf("Exclusions: %d features dropped", nrow(exc_dropped_table)))
```

```{r}
exc_dropped_table
```

### Missingness

Threshold: 10%.

```{r}
# Compute missingness on all columns to ensure a complete record
all_cols <- names(df)
missing_vals <- df %>%
  summarise(across(everything(), ~sum(is.na(.x)) / n() * 100)) %>%
  pivot_longer(everything(), names_to = "feature", values_to = "pct")

miss_dropped_table <- missing_vals %>%
  filter(pct > 10) %>%
  mutate(reason = sprintf("%.1f%% missing", pct)) %>%
  select(feature, reason)

print(sprintf("Missingness: %d features dropped", nrow(miss_dropped_table)))
```

```{r}
miss_dropped_table
```

### Variance

Threshold: 0.01 (raw unscaled variance). This is unit-dependent and primarily catches degenerate columns.

```{r}
# Compute variance on all numeric columns
numeric_all <- df %>% select(where(is.numeric)) %>% names()
var_vals <- df %>%
  summarise(across(all_of(numeric_all), ~var(.x, na.rm = TRUE))) %>%
  pivot_longer(everything(), names_to = "feature", values_to = "variance")

var_dropped_table <- var_vals %>%
  filter(variance < 0.01 | is.na(variance)) %>%
  mutate(reason = sprintf("%.4f", variance)) %>%
  select(feature, reason)

print(sprintf("Variance: %d features dropped", nrow(var_dropped_table)))
```

```{r}
var_dropped_table
```

### Final Feature Set

```{r}
# Combine all dropped features
all_dropped <- bind_rows(
  exc_dropped_table,
  miss_dropped_table,
  var_dropped_table
)

candidates <- setdiff(names(df), all_dropped$feature)
print(sprintf("Final feature set: %d features", length(candidates)))
cat(paste(candidates, collapse = "\n"))
```

## 6. Correlation

We report highly correlated numeric features (|r| > 0.9). This section reports pairs but drops nothing.

```{r}
numeric_final <- candidates[sapply(candidates, function(c) is.numeric(df[[c]]))]
cor_matrix <- cor(df %>% select(all_of(numeric_final)), use = "pairwise.complete.obs")

cor_pairs <- as.data.frame(as.table(cor_matrix)) %>%
  filter(Var1 != Var2) %>%
  filter(abs(Freq) > 0.9) %>%
  distinct(Var1, Var2, .keep_all = TRUE) %>%
  rename(feature1 = Var1, feature2 = Var2, r = Freq)
```

```{r}
if(nrow(cor_pairs) > 0) {
  cor_pairs
} else {
  print("No highly correlated pairs (|r| > 0.9) found.")
}
```

## 7. Write outputs

```{r}
# Analytic dataset: Selected features + scaffolding
df_analytic <- df %>%
  select(all_of(candidates), any_of(c("alias_filled", "observation", "progression_to_death")))

nanoparquet::write_parquet(df_analytic, "data/processed/analytic-dataset.parquet")
```

```{r}
# Selected features artifact
feature_artifact <- missing_vals %>%
  left_join(var_vals, by = "feature") %>%
  filter(feature %in% candidates) %>%
  select(feature, missing_pct = pct, variance)

write.csv(feature_artifact, "output/selected-features.csv", row.names = FALSE)
```

```{r}
print(sprintf("Final feature set: %d features", length(candidates)))
cat(paste(candidates, collapse = "\n"))
```

