

# Exploratory Analysis

Exploratory analysis with detailed missingness patterns.


```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(lubridate)
library(nanoparquet)
```

```{r}
df_raw <- nanoparquet::read_parquet("data/processed/combined-dataset.parquet")

df <- df_raw
```

## Types

```{r}
numeric_cols <- df %>%
  select(where(is.numeric)) %>%
  select(-any_of(c("progression_to_death", "observation"))) %>%
  names()

numeric_cols %>%
  length()
```

This flags 23 columns as numeric. We will spot check them to make sure nothing odd happened.

```{r}
df %>%
  select(all_of(numeric_cols)) %>%
  select(1:4)
```

```{r}
df %>%
  select(all_of(numeric_cols)) %>%
  select(5:8)
```

```{r}
df %>%
  select(all_of(numeric_cols)) %>%
  select(9:12)
```

```{r}
df %>%
  select(all_of(numeric_cols)) %>%
  select(13:16)
```

```{r}
df %>%
  select(all_of(numeric_cols)) %>%
  select(17:20)
```

```{r}
df %>%
  select(all_of(numeric_cols)) %>%
  select(21:23)
```

```{r}
non_numeric_cols <- df %>%
  select(-all_of(numeric_cols)) %>%
  names()

length(non_numeric_cols)
```

```{r}
df %>%
  select(all_of(non_numeric_cols)) %>%
  select(1:4)
```

```{r}
df %>%
  select(all_of(non_numeric_cols)) %>%
  select(5:8)
```

```{r}
df %>%
  select(all_of(non_numeric_cols)) %>%
  select(9:12)
```

```{r}
df %>%
  select(all_of(non_numeric_cols)) %>%
  select(13:16)
```

```{r}
df %>%
  select(all_of(non_numeric_cols)) %>%
  select(17:20)
```

```{r}
df %>%
  select(all_of(non_numeric_cols)) %>%
  select(21:24)
```

```{r}
df %>%
  select(all_of(non_numeric_cols)) %>%
  select(25:28)
```

Removing the dates, most of these look like they a read in correctly

```{r}
date_cols <- df %>%
  select(where(is.POSIXct)) %>%
  names()

date_cols
```

We can then get our official categories.

```{r}
categorical_cols <- df %>%
  select(-all_of(numeric_cols), -all_of(date_cols)) %>%
  names()

categorical_cols %>%
  length()
```

```{r}
df %>%
  select(all_of(categorical_cols)) %>%
  select(1:4) %>%
  print(width = Inf)
```

```{r}
df %>%
  select(all_of(categorical_cols)) %>%
  select(5:8) %>%
  print(width = Inf)
```

```{r}
df %>%
  select(all_of(categorical_cols)) %>%
  select(9:12) %>%
  print(width = Inf)
```

```{r}
df %>%
  select(all_of(categorical_cols)) %>%
  select(13:16) %>%
  print(width = Inf)
```

We determine if this covers all our columns.

```{r}
length(c(categorical_cols, numeric_cols, date_cols))
ncol(df)
```

Not quite.

```{r}
df %>%
  select(-all_of(c(categorical_cols, numeric_cols, date_cols)))

to_factor <- c("alias", "alias_filled", "observation", "progression_to_death")
```
```{r}
df <- df %>%
  mutate_at(to_factor, as.factor)
```

```{r}
df %>%
  select(all_of(to_factor)) %>%
  tail()
```


These are ready to add to the categorical list.

```{r}
categorical_cols <- c(categorical_cols, to_factor)
```

Now we need to inspect this "dcd_map_baseline" column.

```{r}
df %>%
  pull(dcd_map_baseline) %>%
  is.na() %>%
  table()
```

It looks like there is nothing here.

```{r}
exclude_cols <- c("dcd_map_baseline")
```

We still need to make sure the types of our categorical variables are typed correctly.

```{r}
df %>%
  select(all_of(categorical_cols)) %>%
  glimpse()
```

```{r}
df <- df %>%
  mutate_at(categorical_cols, as.factor)

df %>%
  select(all_of(categorical_cols)) %>%
  glimpse()
```

The final set can now be constructed:

```{r}
df <- df %>%
  select(all_of(c(numeric_cols, date_cols, categorical_cols)))
```


## 1. Class and patient distributions

Rows per class:

```{r}
df %>%
  count(progression_to_death)
```

Visualize the patients per class and observations per patient.

```{r}
df %>%
  group_by(progression_to_death, alias_filled) %>%
  summarise(obs = n(), .groups = "drop") %>%
  group_by(progression_to_death) %>%
  summarise(
    count = n(),
    avg_obs = mean(obs),
    max_obs = max(obs),
    min_obs = min(obs)
  )
```

We visualize the distribution of observations per patient, stratified by outcome.
This checks if one class is more "dense" than the other, which could lead to temporal leakage.

```{r}
df %>%
  group_by(progression_to_death, alias_filled) %>%
  summarise(obs = n(), .groups = "drop") %>%
  ggplot(aes(x = obs, fill = factor(progression_to_death))) +
  geom_histogram(binwidth = 1, alpha = 0.4, position = "identity", color = "white") +
  theme_minimal() +
  labs(
    title = "Observations per Patient Distribution",
    x = "Number of Observations",
    y = "Patient Count",
    fill = "Outcome (0=Neg, 1=Pos)"
  )
```

## Missingness

We analyze missing count and percent per column.

```{r}
missing_stats <- df %>%
  summarise(across(everything(), ~ sum(is.na(.x)))) %>%
  pivot_longer(everything(), names_to = "feature", values_to = "null_count") %>%
  mutate(null_pct = null_count / nrow(df) * 100) %>%
  arrange(desc(null_pct))

missing_stats
```

A few of these look really high so we will spot-check a few of the these.

```{r}
df$dcd_map_baseline %>%
  head()

df$dcd_map_baseline %>%
  tail()

df$dcd_map_baseline %>%
  table()
```

```{r}
df$apnea %>%
  head()

df$apnea %>%
  tail()

df$apnea %>%
  table()
```

This seems to be in order.

**LOADBEARING** — A column populated in one workbook and empty in the other is a perfect label proxy. This table is where such leaks are first visible.
Consumed by: the missingness filter, and the leak discussion in `00`'s concat provenance.

```{r}
missing_by_class <- df %>%
  group_by(progression_to_death) %>%
  summarise(
    n_obs = n(),
    across(all_of(numeric_cols), ~ sum(is.na(.x))),
  ) %>%
  pivot_longer(-c(progression_to_death, n_obs),
    names_to = "feature",
    values_to = "null_count"
  ) %>%
  mutate(null_pct = null_count / n_obs * 100) %>%
  filter(null_count > 0) %>%
  select(feature, progression_to_death, null_count, null_pct) %>%
  pivot_wider(
    names_from = progression_to_death,
    values_from = c(null_count, null_pct), names_sep = "_"
  )

missing_by_class
```

We will spot check a few of these to be sure:

```{r}
length(which(is.na(df$age)))

# does it add up to the total
103 + 134
```

```{r}
length(which(is.na(df$heparin_dose_units)))
# does it add up to the total
125 + 134
```

We visualize missingness using a heatmap.

```{r}
image(t(is.na(as.matrix(df[numeric_cols]))), col = c("grey95", "black"), axes = FALSE)
```
A real leakage risk is anything where the proportion of missing values is much higher in one class than the other.


```{r}
df_na_frac <- df %>%
  group_by(progression_to_death) %>%
  summarise(across(all_of(numeric_cols), ~ mean(is.na(.x)))) %>%
  # pivot out the do the calculations and then back because it reads cleaner
  pivot_longer(-progression_to_death, names_to = "feature", values_to = "frac_na") %>%
  pivot_wider(
    names_from = progression_to_death, values_from = frac_na,
    names_prefix = "na_"
  ) %>%
  arrange(desc(pmax(na_0, na_1)))
```


## 5. Feature filters

# Exclusions

**LOADBEARING** — These are post-outcome variables; including any of them makes the ML results meaningless.
Consumed by: `03` model pipeline.

```{r}
leak_justifications <- tibble(
  feature = c(
    "dcd_nrp_total_pump_time",
    "extubation_to_perfusion_warm_ischemic_time",
    "tod_to_perfusion",
    "sbp90_to_declaration",
    "warm_ischemic_time_agonal_phase_to_cooling",
    "did_patient_expire_within_timeframe"
  ),
  reason = c(
    "Post-outcome: pump time recorded during DCD/NRP",
    "Post-outcome: time from extubation to perfusion",
    "Post-outcome: time from declaration to perfusion",
    "Post-outcome: time from SBP<90 to declaration",
    "Post-outcome: agonal phase duration",
    "Direct proxy: patient expiration status"
  )
)

id_excludes <- c("alias", "alias_filled", "observation")
all_excludes <- c(leak_justifications$feature, id_excludes)

exc_dropped_table <- tibble(
  feature = all_excludes,
  reason = if_else(all_excludes %in% leak_justifications$feature,
    leak_justifications$reason[match(all_excludes, leak_justifications$feature)],
    "Manual/ID Exclusion"
  )
)
```

```{r}
leak_justifications
```

```{r}
print(sprintf("Exclusions: %d features dropped", nrow(exc_dropped_table)))
```

```{r}
exc_dropped_table
```

### Missingness

Threshold: 10%.

```{r}
# Compute missingness on all columns to ensure a complete record
all_cols <- names(df)
missing_vals <- df %>%
  summarise(across(everything(), ~ sum(is.na(.x)) / n() * 100)) %>%
  pivot_longer(everything(), names_to = "feature", values_to = "pct")

miss_dropped_table <- missing_vals %>%
  filter(pct > 10) %>%
  mutate(reason = sprintf("%.1f%% missing", pct)) %>%
  select(feature, reason)

print(sprintf("Missingness: %d features dropped", nrow(miss_dropped_table)))
```

```{r}
miss_dropped_table
```

### Variance

Threshold: 0.01 (raw unscaled variance). This is unit-dependent and primarily catches degenerate columns.

```{r}
# Compute variance on all numeric columns
numeric_all <- df %>%
  select(where(is.numeric)) %>%
  names()
var_vals <- df %>%
  summarise(across(all_of(numeric_all), ~ var(.x, na.rm = TRUE))) %>%
  pivot_longer(everything(), names_to = "feature", values_to = "variance")

var_dropped_table <- var_vals %>%
  filter(variance < 0.01 | is.na(variance)) %>%
  mutate(reason = sprintf("%.4f", variance)) %>%
  select(feature, reason)

print(sprintf("Variance: %d features dropped", nrow(var_dropped_table)))
```

```{r}
var_dropped_table
```

### Final Feature Set

```{r}
# Combine all dropped features
all_dropped <- bind_rows(
  exc_dropped_table,
  miss_dropped_table,
  var_dropped_table
)

candidates <- setdiff(names(df), all_dropped$feature)
print(sprintf("Final feature set: %d features", length(candidates)))
cat(paste(candidates, collapse = "\n"))
```

## 4. Descriptive separation by outcome

We check if modeling results are visible in raw data using **first observations only** (one row per patient).

```{r}
df_first <- df %>%
  filter(observation == 1)
```

### Numeric separation (Wilcoxon rank-sum).

```{r}
# Use candidates instead of all numeric columns to ensure leak-free separation analysis
numeric_candidates <- candidates[sapply(candidates, function(c) is.numeric(df[[c]]))]

numeric_sep <- df_first %>%
  select(progression_to_death, all_of(numeric_candidates)) %>%
  pivot_longer(-progression_to_death, names_to = "feature") %>%
  group_by(feature) %>%
  summarise(
    pos_median = median(value[progression_to_death == 1], na.rm = TRUE),
    neg_median = median(value[progression_to_death == 0], na.rm = TRUE),
    p_value = tryCatch(
      wilcox.test(value ~ progression_to_death)$p.value,
      error = function(e) NA_real_
    )
  ) %>%
  arrange(p_value)

numeric_sep %>%
    print(n=Inf)
```

```{r}
sig_numeric <- numeric_sep %>%
  filter(p_value < 0.05) %>%
  pull(feature)

df_first %>%
  select(progression_to_death, all_of(sig_numeric)) %>%
  pivot_longer(-progression_to_death, names_to = "feature", values_to = "value") %>%
  ggplot(aes(x = factor(progression_to_death), y = value, fill = factor(progression_to_death))) +
  geom_boxplot(outlier.alpha = 0.3) +
  facet_wrap(~feature, scales = "free") +
  theme_minimal() +
  labs(
    title = "Significant Numeric Features (p < 0.05)",
    x = "Outcome (0=Neg, 1=Pos)",
    y = "Value",
    fill = "Outcome"
  ) +
  theme(legend.position = "none")
```

```{r}
sig_numeric <- numeric_sep %>%
  filter(p_value < 0.05) %>%
  slice_head(n = 12) %>%
  pull(feature)

df_first %>%
  select(progression_to_death, all_of(sig_numeric)) %>%
  pivot_longer(-progression_to_death, names_to = "feature", values_to = "value") %>%
  ggplot(aes(x = factor(progression_to_death), y = value, fill = factor(progression_to_death))) +
  geom_boxplot(outlier.alpha = 0.3) +
  facet_wrap(~feature, scales = "free") +
  theme_minimal() +
  labs(
    title = "Top Significant Numeric Features (p < 0.05)",
    x = "Outcome (0=Neg, 1=Pos)",
    y = "Value",
    fill = "Outcome"
  ) +
  theme(legend.position = "none")
```

### Categorical separation (Fisher/Chi-sq).

```{r}
# Use candidates instead of all categorical columns
cat_candidates <- candidates[sapply(candidates, function(c) !is.numeric(df[[c]]))]

cat_sep <- df_first %>%
  select(progression_to_death, all_of(cat_candidates)) %>%
  mutate(across(everything(), as.character)) %>%
  pivot_longer(-progression_to_death, names_to = "feature") %>%
  group_by(feature) %>%
  summarise(
    n_levels = n_distinct(value, na.rm = TRUE),
    p_value = tryCatch(
      chisq.test(table(value, progression_to_death))$p.value,
      error = function(e) NA_real_
    )
  ) %>%
  arrange(p_value)

cat_sep %>%
    print(n=Inf)
```

```{r}
sig_categorical <- cat_sep %>%
  filter(p_value < 0.05) %>%
  pull(feature)

df_first %>%
  select(progression_to_death, all_of(sig_categorical)) %>%
  pivot_longer(-progression_to_death, names_to = "feature", values_to = "value") %>%
  ggplot(aes(x = value, fill = factor(progression_to_death))) +
  geom_bar(position = "dodge") +
  facet_wrap(~feature, scales = "free") +
  theme_minimal() +
  labs(
    title = "Significant Categorical Features (p < 0.05)",
    x = "Level",
    y = "Count",
    fill = "Outcome"
  )
```

```{r}
sig_categorical <- cat_sep %>%
  filter(p_value < 0.05) %>%
  slice_head(n = 12) %>%
  pull(feature)

df_first %>%
  select(progression_to_death, all_of(sig_categorical)) %>%
  pivot_longer(-progression_to_death, names_to = "feature", values_to = "value") %>%
  ggplot(aes(x = value, fill = factor(progression_to_death))) +
  geom_bar(position = "dodge") +
  facet_wrap(~feature, scales = "free") +
  theme_minimal() +
  labs(
    title = "Top Significant Categorical Features (p < 0.05)",
    x = "Level",
    y = "Count",
    fill = "Outcome"
  )
```

## 6. Correlation

We report highly correlated numeric features (|r| > 0.9). This section reports pairs but drops nothing.

```{r}
numeric_final <- candidates[sapply(candidates, function(c) is.numeric(df[[c]]))]
cor_matrix <- cor(df %>% select(all_of(numeric_final)), use = "pairwise.complete.obs")

cor_pairs <- as.data.frame(as.table(cor_matrix)) %>%
  filter(Var1 != Var2) %>%
  filter(abs(Freq) > 0.9) %>%
  distinct(Var1, Var2, .keep_all = TRUE) %>%
  rename(feature1 = Var1, feature2 = Var2, r = Freq)
```

```{r}
if (nrow(cor_pairs) > 0) {
  cor_pairs
} else {
  print("No highly correlated pairs (|r| > 0.9) found.")
}
```

## 7. Write outputs

Factors were useful for exploration but are not portable to the Python side, so we cast back before writing: the label and observation index must be integers. Everything else is a character for OneHotEncoder.

```{r}
# as.integer() on a factor returns the level index, not the value, so
# as.character() first.
df_analytic <- df %>%
  select(
    all_of(candidates),
    any_of(c("alias_filled", "observation", "progression_to_death"))
  ) %>%
  mutate(
    progression_to_death = as.integer(as.character(progression_to_death)),
    observation = as.integer(as.character(observation)),
    across(where(is.factor), as.character)
  )

stopifnot(!any(sapply(df_analytic, is.factor)))

nanoparquet::write_parquet(df_analytic, "data/processed/analytic-dataset.parquet")
```

```{r}
# Selected features artifact
feature_artifact <- missing_vals %>%
  left_join(var_vals, by = "feature") %>%
  filter(feature %in% candidates) %>%
  select(feature, missing_pct = pct, variance)

write.csv(feature_artifact, "output/selected-features.csv", row.names = FALSE)
```

```{r}
print(sprintf("Final feature set: %d features", length(candidates)))
cat(paste(candidates, collapse = "\n"))
```
