---
title: 'Statistical Analysis'
---

# Statistical Analysis

## Statistical Framework: Patient-Level GLM
The outcome (progression to death) is a patient-level constant; it does not vary across observations for a single individual. Consequently, a mixed-effects model (GLMM) with a patient random intercept is not identifiable, as there is no within-patient outcome variance to explain.

The **patient-level logistic regression (GLM)** is the proper inferential model for this data structure. By averaging observations per patient, we reduce the problem to a standard binary classification task, avoiding the boundary variance estimates and attenuated fixed effects typical of inappropriately specified mixed models on time-invariant outcomes.

Missing values are handled using Multiple Imputation by Chained Equations (MICE)
with $m=5$ imputations. Each feature is fitted once per imputed dataset, and the
results are pooled using Rubin's rules. This allows the final confidence intervals
to account for the uncertainty introduced by the imputation process.

Both the "all observations" and "first-look only" analyses are pooled across
imputations for consistency. Features are scaled, so estimates are per standard deviation.

**Note on Interpretation:** The `estimate` in the output CSV is the log-odds per standard deviation of the patient mean. These are unadjusted marginal effects. In `04_analyze-model.py`, these are exponentiated to produce Odds Ratios (OR).


```{r}
library(tidyverse)
library(mice)
library(nanoparquet)
```

```{r}
label <- "progression_to_death"
id <- "alias_filled"
m <- 5
dir.create("output", showWarnings = FALSE)

df <- nanoparquet::read_parquet("data/processed/analytic-dataset.parquet")
```

```{r}
print(sprintf("%d rows, %d patients", nrow(df), n_distinct(df[[id]])))
```

## Patient-Level Frames
Average to one row per patient, drop the id, keep only columns with at least two
distinct values, then scale. The distinct-value filter matters: a constant column
becomes all `NaN` under `scale()` and makes mice fail with an unhelpful error.

```{r}
to_patient <- function(d, name = "unnamed") {
  d <- d %>%
    select(-any_of("observation")) %>%
    group_by(.data[[id]], .data[[label]]) %>%
    summarise(
      across(where(is.numeric), ~ mean(.x, na.rm = TRUE)),
      across(where(is.character), ~ first(.x)),
      .groups = "drop"
    ) %>%
    select(-all_of(id))

  cols_before <- names(d)
  d <- d[sapply(d, function(x) n_distinct(x, na.rm = TRUE) >= 2)]
  cols_after <- names(d)
  dropped <- setdiff(cols_before, cols_after)

  if (length(dropped) > 0) {
    print(sprintf(
      "[%s] Dropped %d constant features: %s",
      name, length(dropped), paste(dropped, collapse = ", ")
    ))
  }

  d %>% mutate(across(where(is.numeric) & -all_of(label), ~ as.numeric(scale(.x))))
}
```

```{r}
# We defer calling to_patient until we have the selected feature set from 01_explore-data.qmd
```

```{r}
print(sprintf("%d rows, %d patients", nrow(df), n_distinct(df[[id]])))
```

## Fit
Two loops: features outside, imputations inside, because a feature needs one fit per
imputed dataset before those fits can be pooled. A fitted probability pinned at 0 or 1
means the fit separated, which glm reports as a warning rather than an error, so it is
checked directly and flagged.

```{r}
# Feature Reconciliation: Filter input to only use selected features from 01
selected_feats_csv <- read_csv("output/selected-features.csv", show_col_types = FALSE)
selected_feats <- selected_feats_csv$feature

df_filtered <- df %>% select(all_of(selected_feats), all_of(label), all_of(id), any_of("observation"))
all_obs <- to_patient(df_filtered, "All Observations")
first_obs <- to_patient(df_filtered %>% filter(observation == 1), "First Look Only")

r_feats <- setdiff(names(all_obs), label)
sym_diff <- union(setdiff(r_feats, selected_feats), setdiff(selected_feats, r_feats))

if (length(sym_diff) > 0) {
  print("Note: Some selected features were dropped because they became constant at patient-level:")
  print(setdiff(selected_feats, r_feats))
}
```

```{r}
analyze <- function(d) {
  # **LOADBEARING** — Multiple Imputation (m=5) accounts for missing data uncertainty.
  # Seed is pinned to match ML pipeline for reproducibility.
  # Consumed by: `data/processed/feature_analysis.csv`
  imp <- mice(d, m = m, printFlag = FALSE, seed = 8675309)

  # Imputation Provenance: Log events (collinear columns, etc.)
  # If loggedEvents > 0, mice stripped columns; results may be biased.
  print(sprintf("mice logged events: %d", NROW(imp$loggedEvents)))
  if (NROW(imp$loggedEvents) > 0) {
    print("mice stripped columns due to collinearity/constant values:")
    print(imp$loggedEvents)
  }

  out <- NULL
  for (f in setdiff(names(d), label)) {
    fits <- list()
    separated <- FALSE
    for (i in 1:m) {
      fits[[i]] <- glm(as.formula(sprintf("%s ~ `%s`", label, f)),
        complete(imp, i),
        family = binomial
      )
      p <- fitted(fits[[i]])
      separated <- separated || any(p < 1e-8 | p > 1 - 1e-8)
    }
    # Apply Rubin's Rules via pool()
    s <- summary(pool(as.mira(fits)), conf.int = TRUE)
    s <- s[s$term != "(Intercept)", ]

    # Handle categorical features (multiple levels) by taking the first non-intercept estimate
    # for the purpose of the univariate stability flag.
    est <- s$estimate[1]
    se <- s$std.error[1]
    pval <- s$p.value[1]
    lo <- s[["2.5 %"]][1]
    hi <- s[["97.5 %"]][1]

    out <- rbind(out, data.frame(
      feature = f, estimate = est, std_error = se, p_value = pval,
      ci_low = lo, ci_high = hi,
      unstable = separated || abs(est) > 10 || se > 5,
      separated_flag = separated
    ))
  }
  out %>%
    mutate(p_adj = p.adjust(p_value, "fdr")) %>%
    arrange(p_value)
}
```

```{r}
results <- analyze(all_obs)
first_results <- analyze(first_obs)
```

# Unstable Fits Receipt
unstable_results <- results %>% filter(unstable)
print("\n--- Unstable Fits Receipt ---")
if (nrow(unstable_results) > 0) {
    print(unstable_results %>% select(feature, estimate, std_error, separated_flag))
} else {
    print("No unstable fits detected.")
}
```

```{r}
print(sprintf(
  "Significant (FDR, stable fits): %d of %d. Unstable: %d.",
  sum(results$p_adj < 0.05 & !results$unstable),
  nrow(results), sum(results$unstable)
))
```

```{r}
print(head(results, 15))
```

```{r}
write_csv(results, "data/processed/feature_analysis.csv")
write_csv(first_results, "data/processed/first_look_analysis.csv")
```

## Forest Plot
Odds ratios per SD. Unstable fits are excluded: their intervals span orders of
magnitude and would flatten everything else on a log axis.

```{r}
plot_df <- results %>%
  filter(!unstable) %>%
  mutate(or = exp(estimate), lo = exp(ci_low), hi = exp(ci_high), sig = p_adj < 0.05) %>%
  arrange(or) %>%
  mutate(feature = factor(feature, feature))

ggplot(plot_df, aes(or, feature)) +
  geom_vline(xintercept = 1, color = "red", linetype = "dashed") +
  geom_errorbarh(aes(xmin = lo, xmax = hi), height = 0.2) +
  geom_point(aes(color = sig), size = 2) +
  scale_x_log10() +
  scale_color_manual(values = c("grey70", "firebrick"), labels = c("ns", "FDR p < 0.05"), name = NULL) +
  labs(
    title = "DCD Progression: Unadjusted Marginal Odds Ratios (95% CI)",
    subtitle = sprintf(
      "%d features; %d unstable fits excluded",
      nrow(plot_df), sum(results$unstable)
    ),
    x = "Unadjusted Marginal Odds Ratio per SD (log scale)", y = NULL
  ) +
  theme_minimal()

ggsave("output/glmm_forest_plot_r.png", width = 10, height = 12, dpi = 300)
```

## Limitations

- **Multiple Testing.** We apply FDR correction (Benjamini-Hochberg) across all univariate fits.
  The resulting p_adj indicates the proportion of false discoveries expected among
  significant features.
- **Inferential Nature.** Pooled Odds Ratios are unadjusted. They identify strong candidate
  drivers but do not account for confounders.
- **No mixed-effects model.** The spec called for lme4 with person-level random
  effects, but the outcome is time-invariant within patient, so a random intercept is
  not identifiable. A mixed model here gives boundary variance estimates and
  attenuated fixed effects; the patient-level GLM is the right model for this outcome.
- **No time-to-event analysis.** Only whether the event occurred is recorded, not when,
  so discrete-time survival with a patient frailty term is unavailable.
- **Univariate screens.** Each feature is modelled alone with FDR correction across
  features. The odds ratios are unadjusted, not independent effects.
- **Equal patient weighting.** Averaging gives a patient with twelve observations the
  same weight as one with two.
- **Imputation ignores clustering** and uses every other column as a predictor. Check
  the logged-events count above if the patient count is small.
- **Feature correlation** is covered in `01_explore-data.py`; not duplicated here.
- **Pin the mice version.** `pool` summary column names have changed across releases.

```{r}
print("Analysis complete.")
```
