---
title: 'Statistical Analysis'
---

# Statistical Analysis

## Statistical Framework: First-Look GLM
The outcome (progression to death) is a patient-level constant. The **First-Look logistic regression (GLM)** is the proper inferential model; we analyze only the baseline (first) observation per patient to ensure features precede the outcome and align with the ML unit of analysis.

Missing values are handled using Multiple Imputation by Chained Equations (MICE) with $m=5$ imputations. Each feature is fitted once per imputed dataset, and the results are pooled using Rubin's rules to account for the uncertainty introduced by the imputation process. Features are scaled, so estimates are log-odds per standard deviation.

**Note on Interpretation:** The `estimate` in the output CSV is the log-odds per standard deviation of the baseline measurement (unadjusted marginal effect). These are univariate screens, not mutually adjusted effects. In `04_analyze-model.qmd`, these are exponentiated to produce Odds Ratios (OR).

```{r}
library(tidyverse)
library(mice)
library(nanoparquet)

label <- "progression_to_death"
id <- "alias_filled"
m <- 5
dir.create("output", showWarnings = FALSE)
df <- nanoparquet::read_parquet("data/processed/analytic-dataset.parquet")
```

## Fit
We filter to the first observation, drop constant features to prevent MICE failure, and scale.

```{r}
df_first_look <- df %>% 
  filter(observation == 1) %>% 
  select(-all_of(id), -any_of("observation"))

# Drop constant features
df_first_look <- df_first_look[sapply(df_first_look, function(x) n_distinct(x, na.rm = TRUE) >= 2)]
print(sprintf("Dropped %d constant features", length(names(df)) - ncol(df_first_look)))

# Scale
df_first_look <- df_first_look %>% 
  mutate(across(where(is.numeric) & !all_of(label), ~ as.numeric(scale(.x))))

# MICE Imputation
imp <- mice(df_first_look, m = m, printFlag = FALSE, seed = 8675309)
print(sprintf("mice logged events: %d", NROW(imp$loggedEvents)))

# Pool univariate fits across imputations
results <- map_df(setdiff(names(df_first_look), label), function(f) {
  fits <- lapply(1:m, function(i) {
    glm(as.formula(sprintf("%s ~ `%s`", label, f)), complete(imp, i), family = binomial)
  })
  summary(pool(as.mira(fits)), conf.int = TRUE) %>% 
    filter(term != "(Intercept)") %>% 
    mutate(feature = f) %>% 
    select(feature, estimate, std_error, p_value, ci_low = `2.5 %`, ci_high = `97.5 %`)
}) %>% 
  mutate(p_adj = p.adjust(p_value, "fdr")) %>% 
  arrange(p_value)

write_csv(results, "data/processed/first_look_analysis.csv")
print(head(results, 15))
```

## Forest Plot
Odds ratios per SD.

```{r}
plot_df <- results %>%
  mutate(or = exp(estimate), lo = exp(ci_low), hi = exp(ci_high), sig = p_adj < 0.05) %>%
  arrange(or) %>%
  mutate(feature = factor(feature, feature))

ggplot(plot_df, aes(or, feature)) +
  geom_vline(xintercept = 1, color = "red", linetype = "dashed") +
  geom_errorbarh(aes(xmin = lo, xmax = hi), height = 0.2) +
  geom_point(aes(color = sig), size = 2) +
  scale_x_log10() +
  scale_color_manual(values = c("grey70", "firebrick"), labels = c("ns", "FDR p < 0.05"), name = NULL) +
  labs(
    title = "DCD Progression: Unadjusted Marginal Odds Ratios (95% CI)",
    subtitle = sprintf("%d features; MICE m=5, Rubin's Rules", nrow(plot_df)),
    x = "Unadjusted Marginal Odds Ratio per SD (log scale)", y = NULL
  ) +
  theme_minimal()

ggsave("output/first_look_forest_plot.png", width = 10, height = 12, dpi = 300)
```

## Limitations

- **Scope**: Restricted to baseline (First-Look) observations to prevent leakage; ignores later predictive signals.
- **Unadjusted**: Univariate screens identify candidate drivers but do not account for confounders.
- **Testing**: FDR correction (Benjamini-Hochberg) applied across all univariate fits.
- **Structure**: Imputation ignores clustering; results depend on MICE version for `pool()` summary.

```{r}
print("Analysis complete.")
```
